Learning growth mechanisms of tail realistic preferential attachment models from network degree distributions

### 2025-09-24 Files/paper.tex
Fixing references

Amid the recent rise of interest in networks, there has been a debate on
whether most real networks are scale-free. Claiming a real network is
scale-free is equivalent to saying that its degree distribution follows
a power law, that is the fraction of nodes with degree \(k\) is
proportional to \(k^{-\alpha}\), and therefore has a regularly varying
tail with tail index \(\alpha\). On the side against the claim is
\citet{Broido_2019} who compared the fits of a power law model against
that of several non-scale-free models to nearly a thousand networks,
only to find strong evidence for scale-freeness in four percent and weak
evidence in over half of the networks, thus claiming that scale-free
networks do not make up a majority in real networks. On the other side
of this debate is \citet{Voitalov_2019} who disagrees and claims that
these networks are not nearly as rare and only appear to so as a result
of an unrealistic expectation of a power law without deviations or
noise. Additional evidence of these deviations from a power law is shown
by \citet{Lee24} who demonstrate that a lot of networks are partially
scale-free, in that the body of the degree distribution is often
modelled well by a power law, while the tail is lighter than what is
implied by the body, albeit still regularly varying. Nevertheless, most
studies into the appropriateness of a power law for the degrees of real
networks, the aforementioned references included, have been largely
descriptive in the sense that no information about the growth of the
networks is revealed.

@@

The model from \citet{Barabasi99} provided the foundations for various
generalisations --- \citet{krapivsky01} considered
\(b(k) = (k+1)^\alpha\), and showed that the degree distribution is not
regularly varying (and therefore not following the power law) when
\(0<\alpha<1\), and when \(\alpha>1\) a finite number of nodes end up
with all edges after a certain point resulting in a degenerate degree
distribution. \citet{wang2022random} returns to a linear preference
function of the form \(b(k) = k+\varepsilon\) but adds the possibility
for reciprocal edges to be sent, resulting in the joint distribution of
in-degree and out-degree being multivariate regularly varying and having
the property of hidden regular variation. \citet{rudas07} follows in the
footsteps of \citet{krapivsky01}, by considering a PA tree and using
theory from continuous branching processes, derives a limiting degree
distribution in terms of the preference function \(b(\cdot)\).
Nevertheless, research in this area tends to only focus on the
theoretical asymptotic results of network growth models with little
analysis of real networks.

@@

As the degrees of networks are discrete values, we will use methods from
discrete extreme value theory, in particular those by \citet{shimura12},
who provided theoretical guarantees for a discrete distribution to be
regularly varying or not. Using these results, we demonstrate how the
tail of the degree distribution is affected by \(b(\cdot)\), and
subsequently propose a class of preference functions that is tail
realistic for real networks. These analytical results enables the
likelihood of the degree distribution to be expressed in terms of the
parameters of \(b(\cdot)\), which in turn allows the underlying
mechanism of the network, assumed to grow according to PA, to be
inferred directly.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the PA model alongside the theoretical results
for the survival function of the limiting degree distribution, with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). A class of asymptotically linear preference functions will
be introduced and shown to guarantee regular variation in the degree
distribution while remaining flexible up until a threshold.
Section~\ref{sec-model} utilises the proposed preference function and
illustrates numerically how the tail index of the degree distribution
varies with the model parameters. The simulation study in
Section~\ref{sec-sim} demonstrates that the parameters can be recovered
from fitting the model to only the degree distribution.
Section~\ref{sec-real} fits the model to some real data and provides
posterior estimates for the preference function. Section~\ref{sec-conc}
provides a discussion of this paper and possible avenues for future
work.

@@

\begin{refremark}
If \(b(k)\rightarrow <\infty\), then \(\Omega(F, k) = 0\).

@@

Let \(\bar F (k)\) bet he limiting survival function of degree in a GPA
network with preference function \(b\) such that
\(b(k)\rightarrow\infty\). Then:

@@

Corollary~\ref{cor-omega2} is a direct consequence of
Proposition~\ref{prp-omega} and the results of \citet{shimura12}, and it
aligns with previous findings in the literature. In particular it is
consistent with \citet{krapivsky01}, who showed that a sub-linear
preference function leads to a light-tailed distributions, since for
\(b(k)=k^\alpha\) with \(\alpha<1\), we have
\(\lim_{k\rightarrow}[b(k+1)-b(k)]=0\). Conversely, it also agrees with
the limiting behavior of BA model, which produces a regularly degree
distribution with tail index 2; for the preference function
\(b(k)=k+\varepsilon\), \(\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=1\),
giving a tail index of \(1/\lambda^*\), in which cases the Malthusian
parameter \(\lambda^*\) is 2.

@@

\[
\lim_{k\rightarrow\infty}\Omega(F,k) = \frac{\beta}{\lambda}.
\]

>>>>

Amid the recent rise of interest in networks, there has been a debate on
whether most real networks are scale-free. Claiming a real network is
scale-free is equivalent to saying that its degree distribution follows
a power law, that is the fraction of nodes with degree \(k\) is
proportional to \(k^{-\alpha}\), and therefore has a regularly varying
tail with tail index \(\alpha\). On the side against the claim is
\citet{Broido_2019} who compared the fits of a power law model against
that of several non-scale-free models to nearly a thousand networks,
only to find strong evidence for scale-freeness in four percent and weak
evidence in over half of the networks, thus claiming that scale-free
networks do not make up a majority in real networks. On the other side
of this debate is \citet{Voitalov_2019} who disagrees and claims that
these networks are not nearly as rare and only appear to be so as a
result of an unrealistic expectation of a power law without deviations
or noise. Additional evidence of these deviations from a power law is
shown by \citet{Lee24} who demonstrate that a lot of networks are
partially scale-free, in that the body of the degree distribution is
often modelled well by a power law, while the tail is lighter than what
is implied by the body, albeit still regularly varying. Nevertheless,
most studies into the appropriateness of a power law for the degrees of
real networks, the aforementioned references included, have been largely
descriptive in the sense that no information about the growth of the
networks is revealed.

@@

The model from \citet{Barabasi99} provided the foundations for various
generalisations --- \citet{krapivsky01} considered
\(b(k) = (k+1)^\alpha\), and showed that the degree distribution is not
regularly varying (and therefore not following the power law) when
\(0<\alpha<1\), and when \(\alpha>1\) a finite number of nodes end up
with all edges after a certain point resulting in a degenerate degree
distribution. \citet{wang2022random} returns to a linear preference
function of the form \(b(k) = k+\varepsilon\) but adds the possibility
for reciprocal edges to be sent, resulting in the joint distribution of
in-degree and out-degree being multivariate regularly varying and having
the property of hidden regular variation. \citet{rudas07} followed in
the footsteps of \citet{krapivsky01}, by considering a PA tree and using
theory from continuous branching processes, deriving a limiting degree
distribution in terms of the preference function \(b(\cdot)\).
Nevertheless, research in this area tends to only focus on the
theoretical asymptotic results of network growth models with little
analysis of real networks.

@@

As the degrees of networks are discrete values, we will use methods from
discrete extreme value theory, in particular those by \citet{shimura12},
who provided theoretical guarantees for a discrete distribution to be
regularly varying or not. Using these results, we demonstrate how the
tail of the degree distribution is affected by \(b(\cdot)\), and
subsequently propose a class of preference functions that is tail
realistic for real networks. These analytical results enable the
likelihood of the degree distribution to be expressed in terms of the
parameters of \(b(\cdot)\), which in turn allows the underlying
mechanism of the network, assumed to grow according to PA, to be
inferred directly.

The remainder of the paper is as follows: Section~\ref{sec-tail} gives a
detailed description of the PA model alongside the theoretical results
for the survival function of the limiting degree distribution, with a
focus on the tail behaviour in terms of the preference function
\(b(\cdot)\). A class of asymptotically linear preference functions will
be introduced and shown to guarantee regular variation in the degree
distribution while remaining flexible up until a threshold.
Section~\ref{sec-model} utilises the proposed preference function and
illustrates numerically how the tail index of the degree distribution
varies with the model parameters. The simulation study in Subsection
\ref{sec-sim} demonstrates that the parameters can be recovered from
fitting the model to only the degree distribution. Subsection
\ref{sec-real} fits the model to some real data and provides posterior
estimates for the preference function. Section~\ref{sec-conc} provides a
discussion of this paper and possible avenues for future work.

@@

\begin{refremark}
If \(\lim_{k\rightarrow\infty}b(k)<\infty\), then \(\Omega(F, k) = 0\).

@@

Let \(\bar F (k)\) be the limiting survival function of degrees in a GPA
network with preference function \(b\) such that
\(b(k)\rightarrow\infty\). Then:

@@

Corollary~\ref{cor-omega2} is a direct consequence of
Proposition~\ref{prp-omega} and the results of \citet{shimura12}, and it
aligns with previous findings in the literature. In particular it is
consistent with \citet{krapivsky01}, who showed that a sub-linear
preference function leads to a light-tailed distributions, since for
\(b(k)=k^\alpha\) with \(\alpha<1\), we have
\(\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=0\). Conversely, it also agrees
with the limiting behavior of BA model, which produces a regularly
degree distribution with tail index 2; for the preference function
\(b(k)=k+\varepsilon\), \(\lim_{k\rightarrow\infty}[b(k+1)-b(k)]=1\),
giving a tail index of \(1/\lambda^*\), in which cases the Malthusian
parameter \(\lambda^*\) is 2.

@@

\[
\lim_{k\rightarrow\infty}\Omega(F,k) = \frac{\beta}{\lambda^*}.
\]
